Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2022-01-18 tensor_clustering/tensor_clustering_ieee/ieee_v3.tex
v3

\documentclass[lettersize,onecolumn,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
%\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}
% updated with editorial comments 8/9/2021

@@

%\usepackage{appendix}
\usepackage{fullpage} 
\usepackage{microtype}
\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

@@

\newcommand{\distgap}{D_{\ell_2}}
\newcommand{\Mat}{\text{Mat}}

@@

\author{ Jiaxin Hu and Miaoyan Wang
%IEEE Publication Technology,~\IEEEmembership{Staff,~IEEE,}
        % <-this % stops a space
%\thanks{This paper was produced by the IEEE Publication Technology Group. They are in Piscataway, NJ.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2021; revised August 16, 2021.}
}

@@

\begin{abstract}
We consider the problem of multiway clustering in the presence of unknown degree heterogeneity. Such data problems arise commonly in applications such as recommendation system, neuroimaging, community detection, and hypergraph partitions in social networks. The allowance of degree heterogeneity provides great flexibility in clustering models, but the extra complexity poses significant challenges in both statistics and computation. Here, we develop a degree-corrected tensor block model with estimation accuracy guarantees. We present the phase transition of clustering performance based on the notion of angle separability, and we characterize three signal-to-noise regimes corresponding to different statistical-computational behaviors. In particular, we demonstrate that an intrinsic statistical-to-computational gap emerges only for tensors of order three or greater. Further, we develop an efficient polynomial-time algorithm that provably achieves exact clustering under mild signal conditions. The efficacy of our procedure is demonstrated through two data applications, one on human brain connectome project, and another on Peru Legislation network datasets. 
\end{abstract}

@@

% {
% \color{red}

% \textbf{Other conference paper to journal}
% \begin{enumerate}
%     \item Garvesh. Add more discussion about Theorem 5 and empirical simulations.
%     %\item Eric Chi. Add one more remakr; One more section about the numerical issue; More comparison with other methods; add more future discussion.
%     \item Anru. Add application in introduction; more remarks for algorithm and theorems; One more section about the application to a specific model; more experiments and discussions.

%     \item Christina. Almost re-write the paper. 
% \end{enumerate}
% }

% {
% \color{blue}

% \subsection{Our contribution.} 
% \begin{itemize}
%     \item We propose a generalized dTBM.
%     \item Phase transition.
%     \item We propose a multiway spherical clustering algorithm.
% \end{itemize}

% \subsection{Related work.} We emphasize the comparisons that set our work apart from earlier literature from two perspectives -- models and algorithms.

% From the model perspective, our proposed model is related, but also distinct to following lines of works.
% \begin{itemize}

%     %\item \textit{Tensor block model.}

%     \item \textit{Degree-corrected block model.}

%     \item \textit{Degree-corrected block model for hypergraphs.}

%     \item \textit{Other higher-order clustering models.}
% \end{itemize}

% From the algorithm perspective, our work is inspired by two key algorithmic techniques.

% \begin{itemize}
%     \item \textit{Spherical clustering.}

%     \item \textit{Global-to-local strategy.}
% \end{itemize}

% \subsection{Binary multiway spherical clustering.}

% \paragraph{Algorithm extension}

% \paragraph{Theoretical guarantee}

% }

\section{Introduction}\label{sec:intro}

%\fixme{Jiaxin}{Change the wording: \emph{individual/block effects} $\rightarrow$ \emph{individual/block parameters}, \emph{degree heretogeneity} $\rightarrow$ \emph{non-equal degrees}.}

\IEEEPARstart{M}{ultiway} arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

%To overcome the high-dimensionality, low-rankness is one of the popular structures utilized in tensor data analysis. Specifically, for an order-$K$ tensor $
%\tY$ with low-rank, we have the following decomposition:
%\begin{equation}\label{eq:low_rank}
 %   \tY = \tS \times_1 \mM_1 \cdots \times_K \mM_K, 
%\end{equation}
%where $\tS$ is the low-dimensional core tensor, $\mM_1, ..., \mM_K$ refer to the factor matrices, and $\times_k$ denotes the tensor-to-matrix product.

This paper studies the problem of multiway clustering based on a data tensor. The goal of multiway clustering is to identify a checkerboard structure from a noisy data tensor. Figure~\ref{fig:intro} illustrates the noisy tensor and the underlying structures discovered by multiway clustering methods. The checkerboard structure serves as a meta tool to many popular structures including the low-rankness~\citep{young2018universality}, latent space models~\citep{wang2018learning}, and isotonic models~\citep{pananjady2020isotonic}. In the hypergraph example, the multiway clustering aims to identify the underlying block partition of nodes based on their higher-order connectivities; therefore, we also refer to the clustering as \emph{higher-order clustering}. The most common model for higher-order clustering is called \emph{tensor block model} (TBM)~\citep{wang2019multiway}, which extends the usual matrix stochastic block model~\citep{abbe2017community} to tensors. 
%The analysis tools developed for matrices,
The matrix analysis tools, however, are sub-optimal for higher-order clustering. Developing tensor tools for solving block models has received increased interest recently~\citep{ wang2019multiway,chi2020provable,han2020exact}.

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual effects apart from the block effects. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. Such degree heterogeneity appears commonly in social networks. {\color{blue} Ignoring the individual heterogeneity may seriously mislead the clustering results.} For example, regular block model fails to model {\color{blue} the member affiliation in} the Karate Club network~\citep{bickel2009nonparametric} without addressing
%accounting for 
degree heterogeneity. 
%\fixme{Jiaxin}{Non-degree model ignoring the individual heterogeneity may seriously mislead the clustering results. Karate Club data analysis shows that clustering without heterogeneity tends to assign membership based on the degree of communication (two clusters for leaders and non-leaders) rather than the affiliation (two clusters for people from different clubs). }

The \emph{degree-corrected tensor block model} (dTBM) has been proposed recently to account for the degree heterogeneity~\citep{ke2019community}. The dTBM combines a higher-order checkerboard structure with degree parameter $\mtheta=(\mtheta(1),\ldots,\mtheta(p))^T$ to allow heterogeneity among $p$ nodes.  Figure~\ref{fig:intro} compares the underlying structures of TBM and dTBM with the same number of communities. The dTBM allows varying values within the same community, thereby allowing a richer structure. {\color{blue} To solve dTBM, we project clustering objects to a unit sphere and consider the angle similarity; detailed algorithms are in Section~\ref{sec:alg}. On one hand, the \textit{spherical} clustering avoids the estimation of nuisance degree heterogeneity. On the other hand, the usage of angle similarity brings new challenges to the derivations of theoretical results, and the polar coordinates based techniques are equipped in the proof.
}

% {
% \color{red}
% We use the idea of \emph{spherical clustering} in initialization and iteration. Using angle-based similarity avoids the estimation of $\mtheta$.  
% }

\begin{figure*}[t]
    \centering
    \includegraphics[width = .9\textwidth]{intro2_arxiv.pdf}
    \caption{Examples for order-3 tensor block model (TBM) with and without degree correction. Both TBM and dTBM have four communities on each mode, while dTBM allows a richer structure with degree heterogeneity.
    %\fixme{Miaoyan}{Use horizontal figure} 
    }
    \label{fig:intro}
\end{figure*}

@@

 \item We develop a general dTBM and establish the identifiability for the uniqueness of clustering using the notion of angle seperability.

        \item  We present the phase transition of clustering performance with respect to three different statistical and computational behaviors.  We characterize, for the first time, the critical signal-to-noise (SNR) thresholds in dTBMs, revealing the intrinsic distinctions among (vector) one-dimensional clustering, (matrix) biclustering, and (tensor) higher-order clustering. Specific SNR thresholds and algorithm behaviours are depicted in  Figure~\ref{fig:phase_axis}.

    \item We provide an angle-based algorithm that achieves exact clustering \emph{in polynomial time} under mild conditions. Simulation and data studies demonstrate the outperformance of our algorithm compared with existing higher-order clustering algorithms. 
\end{itemize}
The last two contributions, to our best knowledge, are new to the literature of dTBMs. %Table~\ref{tab:comp} highlights the improvement of our methods over previous results. Our algorithm addresses higher-order tensor data with degree heterogeneity, allows either binary or continuous entries, and achieves exponentially fast rate in misclustering error.

@@

{\bf Related work.} 
{\color{blue}
Our work is closely related to, but also distinct from several lines of existing research. 
\begin{itemize}
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM has been widely used for matrix clustering problems. The theoretical properties and algorithm performance for block models have been well-studied; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood. Table~\ref{tab:comp} summarizes the most relevant models. Specifically,
    \begin{itemize}
        \item the tensor block model (TBM, \cite{wang2019multiway, han2020exact}), a higher-order extension of SBM, fails to allow degree heterogeneity. The Cartesian coordinates based analysis in \cite{han2020exact} is also non-applicable to handle the extra flexibility brought from heterogeneity. In contrast, our model addresses the degree heterogeneity, and the polar coordinates based tools are adapted for the theoretical analysis;
        \item the hypergraph degree-corrected block model (hDCBM) proposed by \cite{ke2019community, yuan2018testing} accounts for degree heterogeneity. Whereas, the hDCBM is designed only for binary observations, and the proposed spectral algorithm in \cite{ke2019community} achieves a sub-optimal clustering accuracy in higher-order scenarios. In constrast, our model allows discrete and continuous entries, and achieves exponentially fast rate in clustering tasks. More importantly, to our best knowledge, we are the first to provide the statistical and computational limits analyses for the degree-corrected block model in tensor clustering.
    \end{itemize}

    \item \textit{Global-to-local algorithm strategy.} Our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{kolda2009tensor}, and we develop a weighted higher-order initialization that relaxes the eigen-gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate provably achieves the exact clustering within only polynomial-time complexity.

\end{itemize}
}

%We emphasize the comparisons that set our work apart from earlier literature from two perspectives---models and algorithms. From the model perspective, our work extends the previous degree-corrected model from matrices to tensors. There is a huge literature on degree-corrected matrix models; see the review paper~\citep{abbe2017community} and the references therein. The tensor counterparts, however, are relatively less understood. Table~\ref{tab:comp} summarizes  the most relevant models for higher-order clustering. Earlier tensor methods either fail to allow degree heterogeneity~\citep{chien2018community,han2020exact,zhang2021exact,ghoshdastidar2017uniform,kim2018stochastic}, or suffer from sub-optimal misclustering rates~\citep{ke2019community,ghoshdastidar2017consistency,chi2020provable}. In contrast, our method addresses the degree heterogeneity, allows discrete and continuous entries, and achieves exponentially fast rate in clustering tasks.

%From the algorithm perspective, our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{kolda2009tensor}, and we develop a weighted higher-order initialization that relaxes the eigen-gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate provably achieves the exact clustering within only polynomial-time complexity.

% {
% \color{red}

% Our work is closely related to, but also distinct from several lines of existing research:
% \begin{enumerate}

% \item \textbf{Degree-corrected model.}
%     \begin{itemize}
%         \item Matrix. (1) Previous work, survey paper (2) Unique technical challenges in Tensor (3) matrix does not show statistical-computational gap;
%         \item Tensor. (Comparison with Ke) (1) Ke does not provide the SNR limits (2) Ke's algorithm is sub-optimal (3) Ke focuses on Binary data. 
%     \end{itemize}

%     \item \textbf{Higher-order clustering.} (1) Previous work fails to account the heterogeneity, which is not realistic; (2) The extension of theoretical results highly challenging.  Extra nuisance parameter makes it more difficult to establish the statistical and computational limits.

%     \item \textbf{Spherical clustering.}

%     \item \textbf{Global-to-local algorithm.} (1) Matrix global-to-local algorithm (2) Tensor global-to-local algorithm is used for non-degree algorithm. Extra nuisance parameter makes it necessary to consider angle-based similarity; angle-based clustering makes the Cartesian coordinates based analysis non-applicable for our setting. 
% \end{enumerate}

% }

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
\color{blue}
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}&  \cite{ghoshdastidar2017consistency} &\cite{ke2019community} & \textbf{Ours}\\
    \hline
        {\small Allow tensors}& $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        {\small Allow heterogeneity} &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        %Assortative assumption&  $\surd$ & $\times$ & $\times$ & $\times$\\
       %{\small Allow various data} & $\times$ & $\surd$ & $\times$ & $\surd$  &$\surd$\\
        {\small Eigen free clustering} &  $\surd$ & $\surd$ & $\times$ & $\times$ &$\surd$ \\
        %{\small Misclustering rate (for order 2$^*$)}& $\exp(-p)$ & $\exp(-p)$ & $p^{-1}$ & $p^{-1}$& $\exp(-p)$\\
        {\small Misclustering rate (for order $K^*$)}& - & $\exp(-p^{K/2})$ & $p^{-1}$& $p^{-2}$ &  $\exp(-p^{K/2})$\\
        \hline
    \end{tabular}
    }
    \caption{Comparison between previous methods with our method. $^*$We list the result for order-K tensors with $K \geq 3$ and general number of communities $r = \tO(1)$.
   % Our results allow general tensors of arbitrary order $K$; See Theorem~\ref{thm:refinement}.
    }\label{tab:comp}
\end{table*}

{\bf Notation.} We use lower-case letters (e.g., $a,b$) for scalars, lower-case boldface letters (e.g., $\ma,\mtheta$) for vectors, upper-case boldface letters (e.g., $\mX,\mY$) for matrices, and calligraphy letters (e.g., $\tX,\tY$) for tensors of order three or greater. We use $|\cdot|$ for the cardinality of a set and $\ind\{\cdot\}$ for the indicator function. For an integer $p\in\mathbb{N}_{+}$, we use the shorthand $[p]= \offf{1,2,...,p}$. For a length-$p$ vector $\ma$, we use $a(i)\in\mathbb{R}$ to denote the $i$-th entry of $\ma$, and use $\ma_{I}$ to denote the sub-vector by restricting the indices in the set $I\subset [p]$.  We use  $\onorm{\ma}=\sqrt{\sum_{i}a^2(i)}$ to denote the $\ell_2$-norm, $\onorm{\ma}_1=\sum_i |a_i|$ to denote the $\ell_1$ norm of $\ma$. For two vector $\ma, \mb$ of the same dimension, we denote the angle between $\ma, \mb$ by 
\begin{equation}
    \cos \of{\ma, \mb} = \frac{\ang{ \ma, \mb}}{ \onorm{\ma} \onorm{\mb} },
\end{equation}
where $\ang{\ma,\mb}$ is the inner product of two vectors and $\cos \of{\ma, \mb} \in [-1,1]$. We make the convention that $\cos \of{\ma, \mb} = \cos \of{\ma^T, \mb^T}$.

For a matrix $\mY$, we use $\mY_{i:}$ to  denote the $i$-th row of the matrix. Let $\tY  \in \bbR^{p_1 \times \cdots \times p_K}$ be an order-$K$ $(p_1,...,p_K)$-dimensional tensor. We use $\tY(i_1,\ldots,i_K)$ to denote the $(i_1,\ldots,i_K)$-th entry of $\tY$. The multilinear multiplication of a tensor $\tS\in \bbR^{r_1\times \cdots \times r_K}$ by matrices $\mM_k \in\mathbb{R}^{p_k\times r_k}$ results in an order-$d$ $(p_1,\ldots,p_K)$-dimensional tensor $\tX$, denoted
\[
\tX=\tS \times_1 \mM_1 \times \cdots \times_K \mM_K,
\]
where the entries of $\tX$ are defined by
\begin{align}
    &\tX(i_1,\ldots, i_K)\\
    &=\sum_{(j_1,\ldots,j_K)}\tS(j_1,\ldots,j_K)\mM_1(i_1,j_1)\cdots \mM_K(i_K,j_K).
\end{align} 
 We use $\text{Ave}(\cdot)$ to denote the operation of taking averages across elements and $\text{Mat}_k(\cdot)$ to denote the unfolding operation that reshapes the tensor along mode $k$ into a matrix. For a symmetric tensor $\tY\in\mathbb{R}^{p\times \cdots \times p}$, we omit the subscript and use $\text{Mat}(\tY)\in\mathbb{R}^{p\times p^{K-1}}$ to denote the unfolding. For two sequences $\{a_p\}, \{b_p\}$, we denote $a_p\lesssim b_p$ if $\lim_{p\to\infty} a_p/b_p \leq c$ and $a_p = \Omega(b_p)$ if $c b_p \leq a_p \leq C b_p$, for some constants $c, C\geq 0$. Throughout the paper, we use the terms ``community'' and ``clusters'' exchangeably. 
 %\fixme{Miaoyan}{\color{red} need to define $\cos(\ma,\mb)$. Be careful with signs.}

{
\color{blue}

{\bf Organization.} The rest of this paper is organized as follows. Section~\ref{sec:model} introduces the degree-corrected tensor block model (dTBM) with three motivating examples and presents the identifiability of dTBM under the angle gap condition. We show the phase transition and the existence of statistical-computational gaps for the higher-order dTBM in Section~\ref{sec:limits}. In Section~\ref{sec:alg}, we provide a polynomial-time two-stage algorithm with misclustering rate guarantees. Numerical studies including the simulations to assess the theoretical results, comparison with other methods, and real data analysis on human brain connectome data and Peru legislation data are in Section~\ref{sec:simulation}. Last, we conclude our paper in Section~\ref{sec:con}.

}

\section{Model formulation}\label{sec:model}

@@

\paragraph{Community detection in hypergraphs} Hypergraph network is a powerful tool to collect the complex entity relations with higher-order interactions~\citep{ke2019community}. A typical undirected hypergraph is denoted as $H = (V,E)$, where $V = [p]$ is the set of nodes and $E$ is the set of undirected hyperedges. Each hyperedge in $E$ is a subset of $V$, and we call the hyperedge an order-$K$ edge if the corresponding subset involves $K$ nodes. We call $H$ a $K$-uniform hypergraph if $E$ only contains order-$K$ edges.

>>>>

\documentclass[lettersize,onecolumn,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}
% updated with editorial comments 8/9/2021

@@

\usepackage{fullpage} 
\usepackage{microtype}
\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

@@

\newcommand{\Mat}{\text{Mat}}

@@

\author{ Jiaxin Hu and Miaoyan Wang\\
University of Wisconsin - Madison
%IEEE Publication Technology,~\IEEEmembership{Staff,~IEEE,}
        % <-this % stops a space
%\thanks{This paper was produced by the IEEE Publication Technology Group. They are in Piscataway, NJ.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2021; revised August 16, 2021.}
}

@@

\begin{abstract}
We consider the problem of multiway clustering in the presence of unknown degree heterogeneity. Such data problems arise commonly in applications such as recommendation system, neuroimaging, community detection, and hypergraph partitions in social networks. The allowance of degree heterogeneity provides great flexibility in clustering models, but the extra complexity poses significant challenges in both statistics and computation. Here, we develop a degree-corrected tensor block model with estimation accuracy guarantees. We present the phase transition of clustering performance based on the notion of angle separability, and we characterize three signal-to-noise regimes corresponding to different statistical-computational behaviors. In particular, we demonstrate that an intrinsic statistical-to-computational gap emerges only for tensors of order three or greater. Further, we develop an efficient polynomial-time algorithm that provably achieves exact clustering under mild signal conditions. The efficacy of our procedure is demonstrated through two data applications, one on human brain connectome project, and another on Peru Legislation network dataset. 
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
\IEEEPARstart{M}{ultiway} arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

We study the problem of multiway clustering based on a data tensor. The goal of multiway clustering is to identify a checkerboard structure from a noisy data tensor. Figure~\ref{fig:intro} illustrates the noisy tensor and the underlying checkerboard structures discovered by multiway clustering methods. In the hypergraph example, the multiway clustering aims to identify the underlying block partition of nodes based on their higher-order connectivities; therefore, we also refer to the clustering as \emph{higher-order clustering}. The most common model for higher-order clustering is called \emph{tensor block model} (TBM)~\citep{wang2019multiway}, which extends the usual matrix stochastic block model~\citep{abbe2017community} to tensors. The matrix analysis tools, however, are sub-optimal for higher-order clustering. Developing tensor tools for solving block models has received increased interest recently~\citep{ wang2019multiway,chi2020provable,han2020exact}.

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual effects apart from the block effects. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. Such degree heterogeneity appears commonly in social networks. Ignoring the degree heterogeneity may seriously mislead the clustering results. For example, regular block model fails to model the member affiliation in the Karate Club network~\citep{bickel2009nonparametric} without addressing degree heterogeneity.

The \emph{degree-corrected tensor block model} (dTBM) has been proposed recently to account for the degree heterogeneity~\citep{ke2019community}. The dTBM combines a higher-order checkerboard structure with degree parameter $\mtheta=(\mtheta(1),\ldots,\mtheta(p))^T$ to allow heterogeneity among $p$ nodes.  Figure~\ref{fig:intro} compares the underlying structures of TBM and dTBM with the same number of communities. The dTBM allows varying values within the same community, thereby allowing a richer structure. To solve dTBM, we project clustering objects to a unit sphere and perform iterative clustering based on angle similarity. We refer to the algorithm as the \textit{spherical} clustering; detailed procedures are in Section~\ref{sec:alg}. The spherical clustering avoids the estimation of nuisance degree heterogeneity. The usage of angle similarity brings new challenges to the theoretical results, and we develop new polar-coordinate based techniques in the proofs.

\begin{figure*}[t]
    \centering
    \includegraphics[width = .9\textwidth]{intro2_arxiv.pdf}
    \caption{Examples for order-3 tensor block model (TBM) with and without degree correction. Both TBM and dTBM have four communities on each mode, while dTBM allows a richer structure with degree heterogeneity.
    }
    \label{fig:intro}
\end{figure*}

@@

 \item We develop a general dTBM and establish the identifiability for the uniqueness of clustering using the notion of angle separability.

\item  We present the phase transition of clustering performance with respect to three different statistical and computational behaviors.  We characterize, for the first time, the critical signal-to-noise (SNR) thresholds in dTBMs, revealing the intrinsic distinctions among (vector) one-dimensional clustering, (matrix) biclustering, and (tensor) higher-order clustering. Specific SNR thresholds and algorithm behaviors are depicted in  Figure~\ref{fig:phase_axis}.

 \item We provide an angle-based algorithm that achieves exact clustering \emph{in polynomial time} under mild conditions. Simulation and data studies demonstrate the outperformance of our algorithm compared with existing higher-order clustering algorithms. 
\end{itemize}
The last two contributions, to our best knowledge, are new to the literature of dTBMs.

@@

{\bf Related work.} 
Our work is closely related to but also distinct from several lines of existing research. Table~\ref{tab:comp} summarizes the most relevant models.

\begin{itemize}
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM have been widely used for matrix clustering problems. The theoretical properties and algorithm performance for matrix block models have been well-studied~\citep{gao2018community}; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood.

    \item  \textit{Tensor block model.} The tensor block model (TBM, \cite{wang2019multiway, han2020exact, ghoshdastidar2017consistency}) is a higher-order extension of SBM, but it fails to allow degree heterogeneity. The Cartesian coordinates based analysis in \cite{han2020exact} is non-applicable to handle the extra complexity brought by degree heterogeneity. In contrast, our model addresses the degree heterogeneity, and we develop polar-coordinate based tools for the theoretical analysis.

    \item \textit{Degree-corrected block model.} The hypergraph degree-corrected block model (hDCBM) proposed by \cite{ke2019community, yuan2018testing} accounts for degree heterogeneity. However, the hDCBM is designed only for binary observations, and the proposed spectral algorithm in \cite{ke2019community} achieves sub-optimal polynomial clustering rate in higher-order scenarios. In constrast, our model allows discrete and continuous entries, and achieves \emph{exponentially} fast rate in clustering tasks. More importantly, to our best knowledge, we are the first to provide the statistical and computational limits analyses for the degree-corrected block model in tensor clustering. See Fig~\ref{fig:phase_axis} for overview of our results.

    \item \textit{Global-to-local algorithm strategy.} Our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{de2000multilinear}, and we develop a weighted higher-order initialization that relaxes the singular-value gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate \emph{provably} achieves the exact clustering within only polynomial-time complexity.

\end{itemize}

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}&  \cite{ghoshdastidar2017consistency} &\cite{ke2019community} & \textbf{Ours}\\
    \hline
      Allow tensors of arbitrary order & $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        Allow degree heterogeneity &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        Singular-value gap-free clustering &  $\surd$ & $\surd$ & $\times$ & $\times$ &$\surd$ \\
        Misclustering rate (for order $K^*$)& - & $\exp(-p^{K/2})$ & $p^{-1}$& $p^{-2}$ &  $\exp(-p^{K/2})$\\
        \hline
    \end{tabular}
    }
    \caption{Comparison between previous methods with our method. $^*$We list the result for order-K tensors with $K \geq 3$ and general number of communities $r = \tO(1)$. }\label{tab:comp}
\end{table*}

{\bf Notation.} We use lower-case letters (e.g., $a,b$) for scalars, lower-case boldface letters (e.g., $\ma,\mtheta$) for vectors, upper-case boldface letters (e.g., $\mX,\mY$) for matrices, and calligraphy letters (e.g., $\tX,\tY$) for tensors of order three or greater. We use $\mone_p$ to denote a vector of length $p$ with all entries to be 1. We use $|\cdot|$ for the cardinality of a set and $\ind\{\cdot\}$ for the indicator function. For an integer $p\in\mathbb{N}_{+}$, we use the shorthand $[p]= \offf{1,2,...,p}$. For a length-$p$ vector $\ma$, we use $a(i)\in\mathbb{R}$ to denote the $i$-th entry of $\ma$, and use $\ma_{I}$ to denote the sub-vector by restricting the indices in the set $I\subset [p]$.  We use  $\onorm{\ma}=\sqrt{\sum_{i}a^2(i)}$ to denote the $\ell_2$-norm, $\onorm{\ma}_1=\sum_i |a_i|$ to denote the $\ell_1$ norm of $\ma$. For two vector $\ma, \mb$ of the same dimension, we denote the angle between $\ma, \mb$ by 
\begin{equation}
    \cos \of{\ma, \mb} = \frac{\ang{ \ma, \mb}}{ \onorm{\ma} \onorm{\mb} },
\end{equation}
where $\ang{\ma,\mb}$ is the inner product of two vectors and $\cos \of{\ma, \mb} \in [-1,1]$. We make the convention that $\cos \of{\ma, \mb} = \cos \of{\ma^T, \mb^T}$.

Let $\tY  \in \bbR^{p_1 \times \cdots \times p_K}$ be an order-$K$ $(p_1,...,p_K)$-dimensional tensor. We use $\tY(i_1,\ldots,i_K)$ to denote the $(i_1,\ldots,i_K)$-th entry of $\tY$. The multilinear multiplication of a tensor $\tS\in \bbR^{r_1\times \cdots \times r_K}$ by matrices $\mM_k \in\mathbb{R}^{p_k\times r_k}$ results in an order-$d$ $(p_1,\ldots,p_K)$-dimensional tensor $\tX$, denoted
\[
\tX=\tS \times_1 \mM_1 \times \cdots \times_K \mM_K,
\]
where the entries of $\tX$ are defined by
\begin{align}
    &\tX(i_1,\ldots, i_K)=\sum_{(j_1,\ldots,j_K)}\tS(j_1,\ldots,j_K)\mM_1(i_1,j_1)\cdots \mM_K(i_K,j_K).
\end{align} 
For a matrix $\mY$, we use $\mY_{i:}$ (respectively, $\mY_{:i}$) to denote the $i$-th row (respectively, $i$-th column) of the matrix. Similarly, for an order-3 tensor, we use $\tY_{::i}$ to denote the $i$-th matrix slide of the tensor. We use $\text{Ave}(\cdot)$ to denote the operation of taking averages across elements and $\text{Mat}_k(\cdot)$ to denote the unfolding operation that reshapes the tensor along mode $k$ into a matrix. For a symmetric tensor $\tY\in\mathbb{R}^{p\times \cdots \times p}$, we omit the subscript and use $\text{Mat}(\tY)\in\mathbb{R}^{p\times p^{K-1}}$ to denote the unfolding. For two sequences $\{a_p\}, \{b_p\}$, we denote $a_p\lesssim b_p$ or $a_p=\tO(b_p)$ if $\lim_{p\to\infty}a_p /b_p\leq c$ for some constant $c\geq 0$, $a_p=o(b_p)$ if $\lim_{p\to\infty}a_p/b_p =0$, and $a_p = \Omega(b_p)$ if both $b_p \lesssim a_p$ and $a_p\lesssim b_p$. Throughout the paper, we use the terms ``community'' and ``clusters'' exchangeably.

{\bf Organization.} The rest of this paper is organized as follows. Section~\ref{sec:model} introduces the degree-corrected tensor block model (dTBM) with three motivating examples and presents the identifiability of dTBM under the angle gap condition. We show the phase transition and the existence of statistical-computational gaps for the higher-order dTBM in Section~\ref{sec:limits}. In Section~\ref{sec:alg}, we provide a polynomial-time two-stage algorithm with misclustering rate guarantees. Numerical studies including the simulation, comparison with other methods, and two real dataset analyses are in Sections~\ref{sec:simulation}-\ref{sec:real}. The main technical ideas we develop for addressing main theorems are provided in Section~\ref{sec:mainproof}. Detailed proofs and extra theoretical results are provided in Appendix.

\section{Model formulation and motivations}\label{sec:model}

@@

\paragraph{Community detection in hypergraphs} Hypergraph network is a powerful tool to represent the complex entity relations with higher-order interactions~\citep{ke2019community}. A typical undirected hypergraph is denoted as $H = (V,E)$, where $V = [p]$ is the set of nodes and $E$ is the set of undirected hyperedges. Each hyperedge in $E$ is a subset of $V$, and we call the hyperedge an order-$K$ edge if the corresponding subset involves $K$ nodes. We call $H$ a $K$-uniform hypergraph if $E$ only contains order-$K$ edges.

It is natural to represent the $K$-uniform hypergraph using a binary order-$K$ adjacency tensor. Let $\tY \in \{0,1\}^{p \times \cdots \times p}$ denote the adjacency tensor, where the entries encode the presence or absence of order-$K$ edges among $p$ nodes. Specifically, for all $(i_1,\ldots,i_K) \in [p]^K$, we have
\begin{equation}
    \tY(i_1,...,i_K) =  \begin{cases}
    1  & \text{if }  (i_1,...,i_K) \in E,\\
    0 & \text{if }  (i_1,...,i_K) \notin E.
    \end{cases}
\end{equation}
